### 2021-11-29 manuscript.tex
Added manuscript file.

\documentclass{article}
\usepackage{authblk}
\usepackage{algorithm}
\usepackage{algorithmic}

\title{Re-initializing k-means clustering with additional observations}
\author[1]{Aaron Lun}
\affil[1]{Genentech, Inc. South San Francisco, CA}

\begin{document}
\maketitle

\section{Motivation}

We consider the problem of k-means clustering after a batch of extra observations have been added to the original dataset.
More specifically, assume that the original dataset has already been subjected to k-means clustering.
After adding the extra observations, we wish to obtain a new clustering of the updated dataset.
This is usually handled by sequential k-means, where the new observations are assigned to (and update) the existing cluster centers;
or by simply repeating the entire k-means procedure on the updated dataset with an initialization step that uses no information from the existing clustering (i.e., a "fresh" clustering).

We would like our new clustering to account for new subpopulations among the recently added observations.
This ensures that our new clustering can capture subpopulations that were not observed in the original dataset,
which is useful when observations are coming in batches from multiple non-replicate sources.
Unfortunately, sequential k-means struggles with this requirement as it starts in a local minimum based on the existing cluster centers.
To escape, a new subpopulation must be large enough to relocate the center of the closest existing cluster, 
while hoping that the observations for that existing cluster can be sensibly reassigned to another existing cluster.

We would also like our new clustering to retain some similarity to the existing clustering of the original dataset.
This is useful when the existing clustering is known to be of interest - 
for example, a domain expert has manually generated annotations for each cluster,
and we want to preserve as much of this information as we can in our new clustering.
It can also be desirable when the existing clustering is high-quality but expensive, e.g., after long iteration times or multiple restarts,
and we would like to maintain quality without repeating the computational work in the new clustering.
A fresh clustering struggles with this requirement as it does not use any information in the existing clustering
and non-obvious partitions may not be recapitulated in the new clustering.

\section{Proposed re-initialization algorithm}

To address these two concerns, we developed a algorithm for ``re-initialization" of a k-means clustering in an updated dataset containing a batch of new observations.
For each existing cluster $i$, we imagine the scenario where that $i$'s center is undefined.
We compute the squared distances for each observation, new and old, to its closest cluster center excluding $i$.
We perform weighted sampling of a single observation where the probabilities are proportional to the squared distances (kmeans++ citation here).
The sampled observation is our proposal for the new center for $i$.
We then assign all observations to their closest clusters (including the proposed new center for $i$) and compute the within-cluster-sum-of-squares (WCSS).
If the WCSS is lower than that with the original center for $i$, we accept this change to $i$'s center; otherwise we continue to use the existing center.
This is repeated for all clusters to obtain an updated set of initial centers that is refined with the usual methods, e.g., Hartigan-Wong or Lloyd.

Algorithm \ref{algo:main} contains a more specific description of the reinitialization procedure.
Here, $\mathcal{C}$ contains the set of $k$ existing centers while $\mathcal{V}$ contains a set of $n$ observations.
We compute the WCSS before ($w$) and after ($\tilde{w}$) the proposed change and accept the proposal if the latter is smaller.
(Note that the centers are used as-is in the WCSS calculation - for simplicity, we do not recompute them from the observations assigned to each cluster.)
We also permit multiple attempts to sample a $\tilde{j}$ that may achieve a lower WCSS.
The algorithm does not make any distinction between the original and new observations - 
both are treated equally, given that the locations of the existing centers are the only information that is retained from the original clustering.

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

\begin{algorithm}
\caption{Pseudocode for k-means reinitialization.}
\begin{algorithmic} 
\REQUIRE $\mathcal{C} = \{\mathbf{c}_1, \mathbf{c}_2, \ldots, \mathbf{c}_k \}$, $\mathcal{V} = \{\mathbf{v}_1, \mathbf{v}_2, \ldots, \mathbf{v}_n \}$
\ENSURE An updated set of centers in $\mathcal{C}$.
\FOR{$i \leftarrow 1$ \TO $k$}
\STATE{
    Define $\mathbf{d} \leftarrow \{d_1, d_2, \ldots, d_n\}$
    \STATE{$w \leftarrow 0$}
    \FOR{$j \leftarrow 1$ \TO $n$}
    \STATE{
        $d_j \leftarrow \min_{\mathbf{c} \in \mathcal{C}} || \mathbf{c} - \mathbf{v}_j ||^2$
        \STATE{$w \leftarrow w + d_j$}
    }
    \ENDFOR

    \FOR{$a \leftarrow 1$ \TO $attempts$}
    \STATE{
        Sample $p$ from $\{1, 2, \ldots, n\}$ with probability vector $\mathbf{d}$
        \STATE{$\tilde{w} \leftarrow 0$}
        \FOR{$j \leftarrow 1$ \TO $n$}
        \STATE{$\tilde{w} \leftarrow \tilde{w} + \min (d_j, || \mathbf{v}_p - \mathbf{v}_j ||^2)$}
        \ENDFOR
        \IF{$w > \tilde{w}$} 
            \STATE{
                $\mathbf{c}_i \leftarrow \mathbf{v}_p$ \\
                \textbf{break}
            }
        \ENDIF
    }
    \ENDFOR
}
\ENDFOR
\end{algorithmic}
\label{algo:main}
\end{algorithm}

The distance-weighted sampling of the reinitialization proposes updates onto observations that are far away from existing centers.
This encourages the creation of new centers inside distinct subpopulations that are not present in the original dataset, allowing us to achieve a large WCSS decrease for the new observations.
We also prefer to relocate centers for weakly separated clusters as their observations can be reassigned to a neighboring cluster without a large WCSS increase.
However, in the absence of any new subpopulations, the existing center is implicitly favored.
Having been through several iterations of refinement during the original clustering, an existing center should achieve a lower WCSS than a relocated center at a random observation.

The number of attempts dictates the effort taken to search for an acceptable update.
A greater number of attempts increases the probability of capturing new subpopulations, as we are more likely to propose centers using observations from those subpopulations.
However, this comes at the cost of some more computational work.
We perform 10 attempts by default, though the exact choice does not seem to have a major impact on the results.

The algorithm runs in $O(nk\log k)$ time where $\log k$ is the time complexity of a vantage point tree-based search (CITATION HERE) to find the closest center.
An implementation can reuse the per-observation identities of the closest centers across iterations to reduce the computational cost.
We can also effectively parallelize the calculations across observations using frameworks like OpenMP.

\section{Evaluation on simulated data}

\section{Evaluation on real data}

\end{document}

### 2021-11-29 manuscript.tex
Added citations, minor rewording here and there.

@@
 
-We consider the problem of k-means clustering after a batch of extra observations have been added to the original dataset.
+We consider the problem of k-means clustering \cite{lloyd1982least} after a batch of extra observations have been added to the original dataset.
 More specifically, assume that the original dataset has already been subjected to k-means clustering.
-After adding the extra observations, we wish to obtain a new clustering of the updated dataset.
-This is usually handled by sequential k-means, where the new observations are assigned to (and update) the existing cluster centers;
-or by simply repeating the entire k-means procedure on the updated dataset with an initialization step that uses no information from the existing clustering (i.e., a "fresh" clustering).
+After adding the extra observations, we wish to obtain a new clustering of the expanded dataset with the same $k$.
+One obvious approach is to just assign each new observation to the closest existing cluster followed by post-initialization refinement - we will refer to this as the simple method.
+The other obvious approach is to repeat the entire k-means procedure on the expanded dataset without using any information from the existing clustering - we will refer to this as a "fresh" clustering.
 
@@
 which is useful when observations are coming in batches from multiple non-replicate sources.
-Unfortunately, sequential k-means struggles with this requirement as it starts in a local minimum based on the existing cluster centers.
+Unfortunately, the simple method struggles with this requirement as it starts in a local minimum based on the existing cluster centers.
 To escape, a new subpopulation must be large enough to relocate the center of the closest existing cluster, 
@@
 It can also be desirable when the existing clustering is high-quality but expensive, e.g., after long iteration times or multiple restarts,
-and we would like to maintain quality without repeating the computational work in the new clustering.
-A fresh clustering struggles with this requirement as it does not use any information in the existing clustering
-and non-obvious partitions may not be recapitulated in the new clustering.
+and we would like to maintain quality without repeating the computational work.
+Such preservation is difficult to achieve in a fresh clustering where non-obvious partitions may not be recapitulated.
 
@@
 
-To address these two concerns, we developed a algorithm for ``re-initialization" of a k-means clustering in an updated dataset containing a batch of new observations.
-For each existing cluster $i$, we imagine the scenario where that $i$'s center is undefined.
-We compute the squared distances for each observation, new and old, to its closest cluster center excluding $i$.
-We perform weighted sampling of a single observation where the probabilities are proportional to the squared distances (kmeans++ citation here).
+To address these two concerns, we developed a algorithm for ``re-initialization" of a k-means clustering in an expanded dataset containing a batch of new observations.
+For each existing cluster $i$, we imagine a scenario where $i$'s center is removed from the set of cluster centers.
+We compute the squared distance of each observation to its closest remaining center.
+We randomly sample a single observation where the sampling probability for each observation is proportional to its squared distance \cite{vassilvitskii2006kmeanspp}.
 The sampled observation is our proposal for the new center for $i$.
-We then assign all observations to their closest clusters (including the proposed new center for $i$) and compute the within-cluster-sum-of-squares (WCSS).
+We then assign all observations to their closest centers (including the proposed new center for $i$) and compute the within-cluster sum-of-squares (WCSS).
 If the WCSS is lower than that with the original center for $i$, we accept this change to $i$'s center; otherwise we continue to use the existing center.
-This is repeated for all clusters to obtain an updated set of initial centers that is refined with the usual methods, e.g., Hartigan-Wong or Lloyd.
+This is repeated for all clusters to obtain an updated set of centers that is refined with the usual methods, e.g., Hartigan-Wong \cite{hartigan1979algorithm} or Lloyd.
 
@@
 (Note that the centers are used as-is in the WCSS calculation - for simplicity, we do not recompute them from the observations assigned to each cluster.)
-We also permit multiple attempts to sample a $\tilde{j}$ that may achieve a lower WCSS.
+We also permit multiple attempts to sample a $p$ that may achieve a lower WCSS.
 The algorithm does not make any distinction between the original and new observations - 
@@
 
-The distance-weighted sampling of the reinitialization proposes updates onto observations that are far away from existing centers.
-This encourages the creation of new centers inside distinct subpopulations that are not present in the original dataset, allowing us to achieve a large WCSS decrease for the new observations.
-We also prefer to relocate centers for weakly separated clusters as their observations can be reassigned to a neighboring cluster without a large WCSS increase.
+The distance-weighted sampling means that new centers are proposed at observations that are far away from existing centers.
+This encourages the creation of new centers inside distinct subpopulations that were not present in the original dataset, allowing us to achieve a large WCSS decrease for the new observations.
+The procedure also prefers to relocate centers for weakly separated clusters as their observations can be reassigned to a neighboring cluster without a large WCSS increase.
 However, in the absence of any new subpopulations, the existing center is implicitly favored.
@@
 
-The algorithm runs in $O(nk\log k)$ time where $\log k$ is the time complexity of a vantage point tree-based search (CITATION HERE) to find the closest center.
+The algorithm runs in $O(nk\log k)$ time where $\log k$ is the time complexity of a vantage point tree-based search \cite{yianilos1993data} to find the closest center.
 An implementation can reuse the per-observation identities of the closest centers across iterations to reduce the computational cost.
@@
 
+\bibliography{ref.bib}
+\bibliographystyle{plain}
+
 \end{document}

### 2021-11-29 package/R/evaluateScenario.R
Compare to a reference method during summarization. Renamed the methods for consistency with the MS.

@@ -13,8 +13,8 @@
 #' The idea is to obtain an initial clustering on \code{x[,subset]} and then apply various strategies to obtain a clustering on \code{x}.
 #' This can be:
 #' \itemize{
-#' \item \code{rest}, where the centers from the initial clustering are used directly for Hartigan-Wong iterations with the full dataset.
-#' \item \code{full}, where a fresh initialization is performed with all observations (not using any information from the initial clustering).
+#' \item \code{simple}, where the centers from the initial clustering are used directly for Hartigan-Wong iterations with the full dataset.
+#' \item \code{fresh}, where a fresh initialization is performed with all observations (not using any information from the initial clustering).
 #' Hartigan-Wong iterations are then performed as usual.
 #' \item \code{reinit}, where a reinitialization strategy is applied that attempts to preserve information from the existing clustering. 
 #' Hartigan-Wong iterations are then performed as usual.
@@ -38,17 +38,17 @@ evaluateScenario <- function(x, subset, k) {
 
     first <- runKmeans(sub, centers=k)
 
-    rest <- runKmeans(x, centers=first$centers)
-    full <- runKmeans(x, centers=k)
+    simple <- runKmeans(x, centers=first$centers)
+    fresh <- runKmeans(x, centers=k)
     reinit <- runKmeans(x, centers=first$centers, reinitialize=TRUE)
 
     data.frame(
-        wcss = c(sum(rest$wcss), sum(full$wcss), sum(reinit$wcss)),
+        wcss = c(sum(simple$wcss), sum(fresh$wcss), sum(reinit$wcss)),
         rand = c(
-            randIndex(first$clusters, rest$clusters[subset]),
-            randIndex(first$clusters, full$clusters[subset]),
+            randIndex(first$clusters, simple$clusters[subset]),
+            randIndex(first$clusters, fresh$clusters[subset]),
             randIndex(first$clusters, reinit$clusters[subset])
         ),
-        method = c("rest", "full", "reinit")
+        method = c("simple", "fresh", "reinit")
     )
 }

### 2021-11-29 package/R/summarizeResults.R

@@ -4,10 +4,14 @@
 #' 
 #' @param df A data frame of WCSS and Rand statistics, usually generated from \code{\link{simulateScenarios}}.
 #' @param collapse Logical scalar indicating whether the results should be collapsed ito a single data frame.
+#' @param ref.method String containing the name of the method to use as the reference.
 #'
 #' @return 
 #' For \code{summarizeResults}, a list of data frames containing the mean and standard error of the Rand index and WCSS for each method.
 #' If \code{collapse=TRUE}, the statistics are rearranged into a single data.frame with a single row. 
+#'
+#' If \code{ref.method} is supplied, the data frames instead contain the mean and standard error of the \emph{ratio} of each method's WCSS to that of the reference method,
+#' and the difference between each method's Rand index and that of the reference method.
 #'
 #' For \code{plotResults}, a list of ggplot objects containing scatter and violin plots.
 #'
@@ -30,13 +34,26 @@
 #' plots$wcss
 #' 
 #' @export
-summarizeResults <- function(df, collapse=FALSE) {
+summarizeResults <- function(df, collapse=FALSE, ref.method=NULL) {
     by.method <- split(seq_len(nrow(df)), df$method)
     wcss <- rand <- list()
-    for (m in names(by.method)) {
-        sub <- df[by.method[[m]],,drop=FALSE]
-        wcss[[m]] <- data.frame(mean = mean(sub$wcss), se = sd(sub$wcss) / sqrt(nrow(sub)))
-        rand[[m]] <- data.frame(mean = mean(sub$rand), se = sd(sub$rand) / sqrt(nrow(sub))) 
+
+    if (is.null(ref.method)) {
+        for (m in names(by.method)) {
+            sub <- df[by.method[[m]],,drop=FALSE]
+            wcss[[m]] <- data.frame(mean = mean(sub$wcss), se = sd(sub$wcss) / sqrt(nrow(sub)))
+            rand[[m]] <- data.frame(mean = mean(sub$rand), se = sd(sub$rand) / sqrt(nrow(sub))) 
+        }
+    } else {
+        refsub <- df[by.method[[ref.method]],,drop=FALSE]
+        for (m in setdiff(names(by.method), ref.method)) {
+            sub <- df[by.method[[m]],,drop=FALSE]
+            stopifnot(nrow(sub) == nrow(refsub))
+            wcss.ratio <- sub$wcss/refsub$wcss
[…]

### 2021-11-29 simulations/populations.R

@@ -31,7 +31,7 @@ for (i in seq_len(nrow(settings))) {
     current <- settings[i,]
     FUN <- generator(npop = current$npop, sd = current$sd, holdout = current$holdout)
     out <- simulateScenario(FUN, k=current$k)
-    collated[[i]] <- summarizeResults(out, collapse=TRUE)
+    collated[[i]] <- summarizeResults(out, collapse=TRUE, ref.method="reinit")
 }
 
 df <- do.call(rbind, collated)
@@ -41,3 +41,6 @@ for (j in seq_len(ncol(df))) {
 
 df <- cbind(settings, df)
 write.table(df, file=file.path(dir, "stats.tsv"), row.names=FALSE, sep="\t")
+
+#hist(log2(df$wcss.simple.mean))
+#hist(df$rand.fresh.mean)

### 2021-11-29 manuscript.tex
Added a table of results for the hold-out simulations.

@@
 
-The algorithm runs in $O(nk\log k)$ time where $\log k$ is the time complexity of a vantage point tree-based search \cite{yianilos1993data} to find the closest center.
+Reinitialization runs in $O(nk\log k)$ time where $\log k$ is the time complexity of a vantage point tree-based search \cite{yianilos1993data} to find the closest center.
 An implementation can reuse the per-observation identities of the closest centers across iterations to reduce the computational cost.
@@
 \section{Evaluation on simulated data}
+
+\subsection{Overview}
+
+Each simulation scenario involves an original dataset plus a new batch of observations.
+We create a clustering on the original dataset by performing kmeans++ initialization \cite{vassilvitskii2006kmeanspp} followed by 10 iterations of Hartigan-Wong refinement.
+We then add the new batch of observations to create the expanded dataset.
+With the reinitialization strategy, we use the method described in Algorithm \ref{algo:main} to update the cluster centers.
+These centers are used in another 10 Hartigan-Wong iterations on the expanded dataset to obtain a new clustering.
+With the simple method, we use the existing cluster centers in the Hartigan-Wong iterations on the expanded dataset without any alteration.
+Finally, with the fresh clustering, we simply repeat the initialization and refinement without using any information from the original clustering.
+
+We use two metrics to evaluate the performance of each strategy.
+We quantify the quality of the clustering by computing the WCSS of the new clustering in the expanded dataset.
+We also compute the adjusted Rand index (ARI) between the original and new clusterings for the subset of observations in the original dataset.
+This captures the similarity in the clusterings after the addition of the new batch.
+
+\subsection{Replicate batches}
+\label{sec:repbatch}
+
+We use a simple simulation framework involving 1000 observations that are randomly distributed among $X$ distinct subpopulations in 5-dimensional space.
+The center $\mathbf{s}_x$ for subpopulation $x$ is sampled from a $\mathcal{N}(\mathbf{0}, \sigma^2 \mathbf{I})$ distribution,
+while the location for each observation in $x$ is sampled from a $\mathcal{N}(\mathbf{s}_x, \mathbf{I})$ distribution.
+We randomly put a proportion $Y$ of observations into the new batch, using the remaining observations as the original dataset.
+In effect, the original dataset and the new batch are replicates of each other as they arise from the same sampling procedure.
+
+\subsection{Batch-specific subpopulations}
+
+We re-use the same framework from Section \ref{sec:repbatch}, but this time, we randomly select $Z$ subpopulations to define the new batch.
+All observations in the selected subpopulations are excluded from the original dataset, meaning that the new batch will contain entirely unique subpopulations.
+
+We observe that the reinitialization strategy regularly achieves the lowest WCSS across scenarios (Table \ref{tab:holdout-subpop}), consistent with its ability to adapt to new subpopulations.
+In fact, the WCSS from reinitialization is often lower than that of a fresh clustering, presumably because the former has a better starting point for refinement.
+The most obvious differences occur when $\sigma$ is large and $Z$ is a large proportion of $X$, such that any failure to form clusters at the new subpopulations inflates the WCSS.
+Reinitialization also achieves higher ARIs compared to a fresh clustering, which is most pronounced when $k$ is larger than $X$;
+this causes the formation of arbitrary partitions within subpopulations that are difficult to recapitulate without any existing information.
+While the ARIs are lower than those of the simple method, this is acceptable given that an existing center needs to be reassigned to a new subpopulation.
+
+\begin{table}
+\caption{Differences in the WCSS and ARI for the simple and fresh clustering strategies compared to reinitialization for a range of simulation scenarios.
+The percentage increase in the WCSS and the absolute increase in the ARI are reported along with the standard error across 100 iterations.}
+\label{tab:holdout-subpop}
+\begin{center}
+\begin{tabular}{r r r r r r r r}
+\hline
+$X$ & $\sigma$ & $Z$ & $k$ & WCSS (simple) & WCSS (fresh) & ARI (simple) & ARI (fresh) \\
+\hline
+5 & 1 & 1 & 5 & $0 \pm 0$ & $0 \pm 0$ & $-0.00 \pm 0.01$ & $-0.06 \pm 0.01$ \\
+10 & 1 & 1 & 5 & $0 \pm 0$ & $0 \pm 0$ & $-0.00 \pm 0.00$ & $-0.14 \pm 0.02$ \\
+20 & 1 & 1 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.27 \pm 0.02$ \\
+5 & 5 & 1 & 5 & $129 \pm 15$ & $53 \pm 11$ & $0.04 \pm 0.01$ & $-0.08 \pm 0.01$ \\
+10 & 5 & 1 & 5 & $6 \pm 1$ & $6 \pm 2$ & $0.07 \pm 0.01$ & $-0.19 \pm 0.02$ \\
+20 & 5 & 1 & 5 & $1 \pm 0$ & $2 \pm 1$ & $0.01 \pm 0.01$ & $-0.34 \pm 0.02$ \\
+5 & 1 & 3 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.01 \pm 0.01$ & $-0.01 \pm 0.01$ \\
+10 & 1 & 3 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.01$ & $-0.07 \pm 0.01$ \\
+20 & 1 & 3 & 5 & $0 \pm 0$ & $0 \pm 0$ & $-0.00 \pm 0.00$ & $-0.20 \pm 0.02$ \\
+5 & 5 & 3 & 5 & $192 \pm 20$ & $38 \pm 10$ & $0.12 \pm 0.01$ & $0.01 \pm 0.01$ \\
+10 & 5 & 3 & 5 & $7 \pm 1$ & $7 \pm 1$ & $0.08 \pm 0.02$ & $-0.11 \pm 0.02$ \\
+20 & 5 & 3 & 5 & $1 \pm 0$ & $3 \pm 1$ & $0.03 \pm 0.01$ & $-0.27 \pm 0.02$ \\
+5 & 1 & 1 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.03 \pm 0.01$ & $-0.21 \pm 0.01$ \\
+10 & 1 & 1 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.27 \pm 0.01$ \\
+20 & 1 & 1 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.32 \pm 0.01$ \\
+5 & 5 & 1 & 10 & $4 \pm 2$ & $0 \pm 0$ & $0.14 \pm 0.01$ & $-0.17 \pm 0.01$ \\
+10 & 5 & 1 & 10 & $58 \pm 5$ & $43 \pm 6$ & $0.06 \pm 0.01$ & $-0.07 \pm 0.01$ \\
+20 & 5 & 1 & 10 & $6 \pm 1$ & $6 \pm 1$ & $0.03 \pm 0.01$ & $-0.24 \pm 0.01$ \\
+5 & 1 & 3 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.05 \pm 0.01$ & $-0.02 \pm 0.01$ \\
+10 & 1 & 3 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.03 \pm 0.01$ & $-0.09 \pm 0.01$ \\
+20 & 1 & 3 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.22 \pm 0.01$ \\
+5 & 5 & 3 & 10 & $45 \pm 11$ & $0 \pm 0$ & $0.26 \pm 0.01$ & $0.03 \pm 0.01$ \\
+10 & 5 & 3 & 10 & $98 \pm 10$ & $31 \pm 4$ & $0.04 \pm 0.01$ & $-0.05 \pm 0.01$ \\
+20 & 5 & 3 & 10 & $9 \pm 1$ & $8 \pm 1$ & $0.06 \pm 0.01$ & $-0.16 \pm 0.01$ \\
+5 & 1 & 1 & 20 & $1 \pm 0$ & $0 \pm 0$ & $0.05 \pm 0.01$ & $-0.24 \pm 0.01$ \\
+10 & 1 & 1 & 20 & $0 \pm 0$ & $0 \pm 0$ & $0.03 \pm 0.01$ & $-0.34 \pm 0.01$ \\
+20 & 1 & 1 & 20 & $0 \pm 0$ & $0 \pm 0$ & $0.01 \pm 0.00$ & $-0.42 \pm 0.01$ \\
+5 & 5 & 1 & 20 & $4 \pm 0$ & $1 \pm 0$ & $0.22 \pm 0.01$ & $-0.17 \pm 0.01$ \\
+10 & 5 & 1 & 20 & $9 \pm 3$ & $1 \pm 0$ & $0.07 \pm 0.01$ & $-0.24 \pm 0.01$ \\
+20 & 5 & 1 & 20 & $37 \pm 3$ & $35 \pm 3$ & $0.08 \pm 0.01$ & $-0.07 \pm 0.01$ \\
+5 & 1 & 3 & 20 & $1 \pm 0$ & $0 \pm 0$ & $0.06 \pm 0.01$ & $-0.02 \pm 0.01$ \\
+10 & 1 & 3 & 20 & $0 \pm 0$ & $0 \pm 0$ & $0.02 \pm 0.01$ & $-0.19 \pm 0.01$ \\
+20 & 1 & 3 & 20 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.01$ & $-0.31 \pm 0.01$ \\
+5 & 5 & 3 & 20 & $31 \pm 9$ & $1 \pm 0$ & $0.36 \pm 0.01$ & $0.04 \pm 0.01$ \\
+10 & 5 & 3 & 20 & $16 \pm 4$ & $1 \pm 0$ & $0.16 \pm 0.01$ & $-0.08 \pm 0.01$ \\
+20 & 5 & 3 & 20 & $48 \pm 3$ & $33 \pm 3$ & $0.06 \pm 0.01$ & $-0.07 \pm 0.01$ \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
 

### 2021-11-29 simulations/reformatter.R

@@ -1,0 +1,9 @@
+df <- read.delim("results/populations/stats.tsv")
+X <- sprintf("%s & %s & %s & %s & $%d \\pm %d$ & $%s \\pm %s$ & $%s \\pm %s$ & $%s \\pm %s$ \\\\",
+    df$npop, df$sd, df$holdout, df$k,
+    round((df$wcss.simple.mean - 1)*100), round((df$wcss.simple.se)*100),
+    round((df$wcss.fresh.mean - 1)*100), round((df$wcss.fresh.se)*100),
+    round(df$rand.simple.mean, 2), round(df$rand.simple.se, 2),
+    round(df$rand.fresh.mean, 2), round(df$rand.fresh.se, 2)
+)
+cat(X, sep="\n")

### 2021-11-29 package/fetch_headers.sh
Fixed the header-fetching script.

@@ -8,13 +8,13 @@ fi
 # Transferring header files.
 cd _cpp
 git checkout b6dbb52353c5d2a2ffaa94ecaf3a811bf4480b31
-cp -r include/kmeans ../src
+cp -r include/kmeans ../inst/include
 
 # Fetching dependencies.
 cmake -B build -S .
-cp -r build/_deps/aarand-src/include/aarand ../src
+cp -r build/_deps/aarand-src/include/aarand ../inst/include
 
 # Committing everything.
 cd ..
-git add src/kmeans
-git add src/aarand
+git add inst/include/kmeans
+git add inst/include/aarand

### 2021-11-30 simulations/circle.R
Added more simulations for pure replicates and circles. Generalized the formatter for other simulation scenarios.

@@ -1,0 +1,47 @@
+generator <- function(sd = 1, exclude = 0.5, ncells = 1000) {
+    force(sd)
+    force(exclude)
+    force(ncells)
+    function() {
+        i <- seq_len(ncells)
+        theta <- i / ncells * 2 * pi
+        x <- sin(theta)
+        y <- cos(theta)
+
+        stuff <- rbind(x, y)
+        stuff <- stuff + rnorm(length(stuff), sd=sd)
+        keep <- i <= (1-exclude) * ncells
+
+        list(data=stuff, subset=keep)
+    }
+}
+
+set.seed(984241)
+
+settings <- expand.grid(
+    sd = c(0.1, 0.5),
+    exclude = c(0.1, 0.25, 0.5),
+    k = c(5, 10)
+)
+
+library(KmeansReinitialization2021)
+collated <- list()
+
+for (i in seq_len(nrow(settings))) {
+    current <- settings[i,]
+    FUN <- generator(exclude = current$exclude, sd = current$sd)
+    out <- simulateScenario(FUN, k=current$k)
+    collated[[i]] <- summarizeResults(out, collapse=TRUE, ref.method="reinit")
+}
+
+df <- do.call(rbind, collated)
+for (j in seq_len(ncol(df))) {
+    df[,j] <- signif(df[,j], 4)
+}
+
+df <- cbind(settings, df)
+dir <- "results"
+write.table(df, file=file.path(dir, "circle.tsv"), row.names=FALSE, sep="\t")
+
+#hist(log2(df$wcss.simple.mean))
+#hist(df$rand.fresh.mean)

### 2021-11-30 simulations/formatter.R

@@ -1,0 +1,19 @@
+path <- tail(commandArgs(), 1)
+df <- read.delim(path)
+collated <- vector("list", ncol(df))
+
+# Assembling statistics.
+X <- sprintf("$%d \\pm %d$ & $%d \\pm %d$ & $%.02f \\pm %.02f$ & $%.02f \\pm %.02f$ \\\\",
+    round((df$wcss.simple.mean - 1)*100), round((df$wcss.simple.se)*100),
+    round((df$wcss.fresh.mean - 1)*100), round((df$wcss.fresh.se)*100),
+    round(df$rand.simple.mean, 2), round(df$rand.simple.se, 2),
+    round(df$rand.fresh.mean, 2), round(df$rand.fresh.se, 2)
+)
+
+# Everything else is a parameter column.
+is.stat <- grepl("^(wcss|rand)", colnames(df))
+collated <- as.list(df[,!is.stat])
+collated <- append(collated, list(X))
+
+output <- do.call(paste, c(collated, sep=" & "))
+cat(output, sep="\n")

### 2021-11-30 simulations/newpop.R

@@ -12,9 +12,6 @@ generator <- function(npop = 10, sd = 5, holdout=1, ncells = 1000, ndim = 5) {
     }
 }
 
-dir <- "results/populations"
-unlink(dir, recursive=TRUE)
-dir.create(dir, recursive=TRUE, showWarnings=FALSE)
 set.seed(984241)
 
 settings <- expand.grid(
@@ -40,7 +37,8 @@ for (j in seq_len(ncol(df))) {
 }
 
 df <- cbind(settings, df)
-write.table(df, file=file.path(dir, "stats.tsv"), row.names=FALSE, sep="\t")
+dir <- "results"
+write.table(df, file=file.path(dir, "newpop.tsv"), row.names=FALSE, sep="\t")
 
 #hist(log2(df$wcss.simple.mean))
 #hist(df$rand.fresh.mean)

### 2021-11-30 simulations/reformatter.R

@@ -1,9 +1,0 @@
-df <- read.delim("results/populations/stats.tsv")
-X <- sprintf("%s & %s & %s & %s & $%d \\pm %d$ & $%s \\pm %s$ & $%s \\pm %s$ & $%s \\pm %s$ \\\\",
-    df$npop, df$sd, df$holdout, df$k,
-    round((df$wcss.simple.mean - 1)*100), round((df$wcss.simple.se)*100),
-    round((df$wcss.fresh.mean - 1)*100), round((df$wcss.fresh.se)*100),
-    round(df$rand.simple.mean, 2), round(df$rand.simple.se, 2),
-    round(df$rand.fresh.mean, 2), round(df$rand.fresh.se, 2)
-)
-cat(X, sep="\n")

### 2021-11-30 simulations/replicate.R

@@ -1,0 +1,47 @@
+generator <- function(npop = 10, sd = 5, holdout=0.5, ncells = 1000, ndim = 5) {
+    force(npop)
+    force(ndim)
+    force(sd)
+    force(ncells)
+    function() {
+        populations <- matrix(rnorm(npop * ndim, sd=sd), ncol=npop)
+        sampled <- sample(ncol(populations), ncells, replace=TRUE)
+        stuff <- populations[,sampled]
+        stuff <- stuff + rnorm(length(stuff))
+
+        keep <- !logical(ncells)
+        keep[sample(ncells, ncells * holdout)] <- FALSE
+        list(data=stuff, subset=keep)
+    }
+}
+
+set.seed(984241)
+
+settings <- expand.grid(
+    npop = c(5, 10),
+    sd = c(1, 5),
+    holdout = c(0.1, 0.5),
+    k = c(5, 10)
+)
+
+library(KmeansReinitialization2021)
+collated <- list()
+
+for (i in seq_len(nrow(settings))) {
+    current <- settings[i,]
+    FUN <- generator(npop = current$npop, sd = current$sd, holdout = current$holdout)
+    out <- simulateScenario(FUN, k=current$k)
+    collated[[i]] <- summarizeResults(out, collapse=TRUE, ref.method="reinit")
+}
+
+df <- do.call(rbind, collated)
+for (j in seq_len(ncol(df))) {
+    df[,j] <- signif(df[,j], 4)
+}
+
+df <- cbind(settings, df)
+dir <- "results"
+write.table(df, file=file.path(dir, "replicate.tsv"), row.names=FALSE, sep="\t")
+
+#hist(log2(df$wcss.simple.mean))
+#hist(df$rand.fresh.mean)

### 2021-11-30 manuscript.tex
Added text for the latest simulations.

@@
 
-Each simulation scenario involves an original dataset plus a new batch of observations.
-We create a clustering on the original dataset by performing kmeans++ initialization \cite{vassilvitskii2006kmeanspp} followed by 10 iterations of Hartigan-Wong refinement.
-We then add the new batch of observations to create the expanded dataset.
-With the reinitialization strategy, we use the method described in Algorithm \ref{algo:main} to update the cluster centers.
+Each simulation scenario involves the creation of one set of observations representing the original dataset and another set representing the new batch.
+We cluster the original dataset by performing kmeans++ initialization \cite{vassilvitskii2006kmeanspp} followed by 10 iterations of Hartigan-Wong refinement.
+We then add the new batch of observations to create the expanded dataset that is subjected to further clustering with some of the previously described methods.
+For reinitialization, we use an implementation of Algorithm \ref{algo:main} to update the cluster centers.
 These centers are used in another 10 Hartigan-Wong iterations on the expanded dataset to obtain a new clustering.
-With the simple method, we use the existing cluster centers in the Hartigan-Wong iterations on the expanded dataset without any alteration.
-Finally, with the fresh clustering, we simply repeat the initialization and refinement without using any information from the original clustering.
+For the simple method, we directly use the existing cluster centers in the Hartigan-Wong iterations on the expanded dataset without any alteration.
+Finally, for the fresh clustering, we repeat the initialization and refinement without using any values from the original clustering.
 
-We use two metrics to evaluate the performance of each strategy.
+We use two metrics to evaluate the performance of each strategy in each simulation scenario.
 We quantify the quality of the clustering by computing the WCSS of the new clustering in the expanded dataset.
 We also compute the adjusted Rand index (ARI) between the original and new clusterings for the subset of observations in the original dataset.
-This captures the similarity in the clusterings after the addition of the new batch.
+This represents the relative similarity in the clusterings after the addition of the new batch.
 
@@
 
-We use a simple simulation framework involving 1000 observations that are randomly distributed among $X$ distinct subpopulations in 5-dimensional space.
+We start with a simple simulation framework involving 1000 observations that are randomly distributed among $X$ distinct subpopulations in 5-dimensional space.
 The center $\mathbf{s}_x$ for subpopulation $x$ is sampled from a $\mathcal{N}(\mathbf{0}, \sigma^2 \mathbf{I})$ distribution,
 while the location for each observation in $x$ is sampled from a $\mathcal{N}(\mathbf{s}_x, \mathbf{I})$ distribution.
-We randomly put a proportion $Y$ of observations into the new batch, using the remaining observations as the original dataset.
+We randomly select a proportion $Y$ of observations and put them into the new batch; the remaining observations are used as the original dataset.
 In effect, the original dataset and the new batch are replicates of each other as they arise from the same sampling procedure.
+
+For these scenarios, the reinitialization strategy achieves the lowest WCSS and close to the highest ARI (Table \ref{tab:replicate}).
+We see the greatest difference in the WCSS when $k$ is close to $X$, likely due to a poor initialization in the original or fresh clusterings;
+in contrast, reinitialization has a chance to overcome a poor start by only accepting updates that decrease the WCSS.
+ARIs are highest for the simple method, consistent with its complete reliance on the existing cluster centers.
+Nonetheless, the ARIs from reinitialization are close to the maximum and are consistently larger than those from a fresh clustering, as expected.
+
+\begin{table}
+\caption{Differences in the WCSS and ARI for the simple and fresh clustering strategies compared to reinitialization for a range of simulations with replicate batches.
+The percentage increase in the WCSS and the absolute increase in the ARI are reported along with the standard error across 100 iterations.}
+\label{tab:replicate}
+\begin{center}
+\begin{tabular}{r r r r r r r r}
+\hline
+$X$ & $\sigma$ & $Y$ & $k$ & WCSS (simple) & WCSS (fresh) & ARI (simple) & ARI (fresh) \\
+\hline
+5 & 1 & 0.1 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.14 \pm 0.02$ \\
+10 & 1 & 0.1 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.20 \pm 0.02$ \\
+5 & 5 & 0.1 & 5 & $32 \pm 8$ & $34 \pm 6$ & $0.05 \pm 0.01$ & $-0.06 \pm 0.01$ \\
+10 & 5 & 0.1 & 5 & $6 \pm 1$ & $6 \pm 2$ & $0.04 \pm 0.01$ & $-0.26 \pm 0.02$ \\
+5 & 1 & 0.5 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.08 \pm 0.01$ \\
+10 & 1 & 0.5 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.15 \pm 0.02$ \\
+5 & 5 & 0.5 & 5 & $42 \pm 10$ & $44 \pm 9$ & $0.07 \pm 0.01$ & $-0.05 \pm 0.01$ \\
+10 & 5 & 0.5 & 5 & $2 \pm 1$ & $4 \pm 1$ & $0.02 \pm 0.01$ & $-0.28 \pm 0.02$ \\
+5 & 1 & 0.1 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.42 \pm 0.01$ \\
+10 & 1 & 0.1 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.31 \pm 0.01$ \\
+5 & 5 & 0.1 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.32 \pm 0.01$ \\
+10 & 5 & 0.1 & 10 & $38 \pm 4$ & $47 \pm 5$ & $0.10 \pm 0.01$ & $-0.06 \pm 0.01$ \\
+5 & 1 & 0.5 & 10 & $0 \pm 0$ & $0 \pm 0$ & $-0.00 \pm 0.00$ & $-0.27 \pm 0.01$ \\
+10 & 1 & 0.5 & 10 & $0 \pm 0$ & $0 \pm 0$ & $-0.00 \pm 0.00$ & $-0.20 \pm 0.01$ \\
+5 & 5 & 0.5 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.01 \pm 0.00$ & $-0.27 \pm 0.01$ \\
+10 & 5 & 0.5 & 10 & $43 \pm 5$ & $34 \pm 4$ & $0.10 \pm 0.01$ & $-0.04 \pm 0.01$ \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
 
@@
 We observe that the reinitialization strategy regularly achieves the lowest WCSS across scenarios (Table \ref{tab:holdout-subpop}), consistent with its ability to adapt to new subpopulations.
-In fact, the WCSS from reinitialization is often lower than that of a fresh clustering, presumably because the former has a better starting point for refinement.
-The most obvious differences occur when $\sigma$ is large and $Z$ is a large proportion of $X$, such that any failure to form clusters at the new subpopulations inflates the WCSS.
-Reinitialization also achieves higher ARIs compared to a fresh clustering, which is most pronounced when $k$ is larger than $X$;
+The most obvious differences occur when $\sigma$ is large and $Z$ is a large proportion of $X$, such that any failure to form distinct clusters at the new subpopulations inflates the WCSS.
+(The effect of $k \approx X$ seen in Section \ref{sec:repbatch} is also relevant here.)
+Reinitialization achieves higher ARIs compared to a fresh clustering, which is most pronounced when $k$ is larger than $X$;
 this causes the formation of arbitrary partitions within subpopulations that are difficult to recapitulate without any existing information.
-While the ARIs are lower than those of the simple method, this is acceptable given that an existing center needs to be reassigned to a new subpopulation.
+While the ARIs are lower than those of the simple method, this is arguably acceptable given that an existing center needs to be reassigned to a new subpopulation.
 
 \begin{table}
-\caption{Differences in the WCSS and ARI for the simple and fresh clustering strategies compared to reinitialization for a range of simulation scenarios.
+\caption{Differences in the WCSS and ARI for the simple and fresh clustering strategies compared to reinitialization for a range of simulations involving batch-specific subpopulations. 
 The percentage increase in the WCSS and the absolute increase in the ARI are reported along with the standard error across 100 iterations.}
@@
 
+\subsection{Continuous trajectories}
+
+Finally, we simulate points lying in a circle on a 2-dimensional plane with random $\mathcal{N}(0, \sigma^2\mathbf{I})$ noise.
+To create the new batch, we take all cells in an arc of the circle corresponding to a proportion $U$ of the circumference.
+This represents a continuous trajectory of new observations rather than discrete subpopulations.
+
+WCSS values are comparable for all methods while the highest ARIs are shared between reinitialization and the simple method (Table \ref{tab:circular}).
+The fresh clustering has the lowest ARI, which is expected given the absence of a clear partitioning in a circular mass of observations.
+The low WCSS for the simple method is due to the proximity of the new and existing observations,
+allowing the centers to incrementally migrate into the new batch during refinement.
+This is more difficult in scenarios involving new subpopulations where the simple method is more likely to be trapped in a local minimum.
+
+\begin{table}
+\caption{Differences in the WCSS and ARI for the simple and fresh clustering strategies compared to reinitialization for a range of simulations involving a circular trajectory.
+The percentage increase in the WCSS and the absolute increase in the ARI are reported along with the standard error across 100 iterations.}
+\label{tab:circular}
+\begin{center}
+\begin{tabular}{r r r r r r r}
+\hline
+$\sigma$ & $A$ & $k$ & WCSS (simple) & WCSS (fresh) & ARI (simple) & ARI (fresh) \\
+\hline
+0.1 & 0.1 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.02 \pm 0.01$ & $-0.17 \pm 0.02$ \\
+0.5 & 0.1 & 5 & $0 \pm 0$ & $0 \pm 0$ & $-0.00 \pm 0.00$ & $-0.21 \pm 0.02$ \\
+0.1 & 0.25 & 5 & $0 \pm 0$ & $0 \pm 0$ & $-0.04 \pm 0.01$ & $-0.02 \pm 0.01$ \\
+0.5 & 0.25 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.01 \pm 0.01$ & $-0.08 \pm 0.01$ \\
+0.1 & 0.5 & 5 & $0 \pm 0$ & $0 \pm 0$ & $-0.01 \pm 0.01$ & $-0.01 \pm 0.01$ \\
+0.5 & 0.5 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.02 \pm 0.01$ & $0.00 \pm 0.01$ \\
+0.1 & 0.1 & 10 & $1 \pm 0$ & $1 \pm 0$ & $0.01 \pm 0.01$ & $-0.10 \pm 0.01$ \\
+0.5 & 0.1 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.01 \pm 0.01$ & $-0.30 \pm 0.01$ \\
+0.1 & 0.25 & 10 & $1 \pm 0$ & $0 \pm 0$ & $-0.01 \pm 0.01$ & $-0.02 \pm 0.01$ \\
+0.5 & 0.25 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.02 \pm 0.01$ & $-0.12 \pm 0.01$ \\
+0.1 & 0.5 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.03 \pm 0.01$ & $0.01 \pm 0.01$ \\
+0.5 & 0.5 & 10 & $0 \pm 0$ & $0 \pm 0$ & $-0.01 \pm 0.01$ & $-0.02 \pm 0.01$ \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
 \section{Evaluation on real data}

### 2021-12-01 manuscript.tex
Reduce the number of simulation scenarios, cleaned up text.

@@
 For these scenarios, the reinitialization strategy achieves the lowest WCSS and close to the highest ARI (Table \ref{tab:replicate}).
-We see the greatest difference in the WCSS when $k$ is close to $X$, likely due to a poor initialization in the original or fresh clusterings;
-in contrast, reinitialization has a chance to overcome a poor start by only accepting updates that decrease the WCSS.
-ARIs are highest for the simple method, consistent with its complete reliance on the existing cluster centers.
+We see the greatest differences in the WCSS when $k$ is close to $X$, likely due to a poor initialization in the original or fresh clusterings;
+in contrast, reinitialization has an opportunity to overcome a poor start by only accepting updates that decrease the WCSS.
+ARIs are highest for the simple method, consistent with its reliance on the existing cluster centers.
 Nonetheless, the ARIs from reinitialization are close to the maximum and are consistently larger than those from a fresh clustering, as expected.
@@
 
-We re-use the same framework from Section \ref{sec:repbatch}, but this time, we randomly select $Z$ subpopulations to define the new batch.
-All observations in the selected subpopulations are excluded from the original dataset, meaning that the new batch will contain entirely unique subpopulations.
+Re-using the same framework from Section \ref{sec:repbatch}, we randomly select $Z$ subpopulations and exclude all of its observations from the original dataset.
+This means that the new batch will contain entirely unique subpopulations.
 
@@
 Reinitialization achieves higher ARIs compared to a fresh clustering, which is most pronounced when $k$ is larger than $X$;
-this causes the formation of arbitrary partitions within subpopulations that are difficult to recapitulate without any existing information.
-While the ARIs are lower than those of the simple method, this is arguably acceptable given that an existing center needs to be reassigned to a new subpopulation.
+this is due to the formation of arbitrary partitions within subpopulations that are difficult to recapitulate without any existing information.
+While the reinitialization ARIs are lower than those of the simple method, this is arguably acceptable given that an existing center needs to be reassigned to a new subpopulation.
 
@@
 \hline
-5 & 1 & 1 & 5 & $0 \pm 0$ & $0 \pm 0$ & $-0.00 \pm 0.01$ & $-0.06 \pm 0.01$ \\
-10 & 1 & 1 & 5 & $0 \pm 0$ & $0 \pm 0$ & $-0.00 \pm 0.00$ & $-0.14 \pm 0.02$ \\
-20 & 1 & 1 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.27 \pm 0.02$ \\
-5 & 5 & 1 & 5 & $129 \pm 15$ & $53 \pm 11$ & $0.04 \pm 0.01$ & $-0.08 \pm 0.01$ \\
-10 & 5 & 1 & 5 & $6 \pm 1$ & $6 \pm 2$ & $0.07 \pm 0.01$ & $-0.19 \pm 0.02$ \\
-20 & 5 & 1 & 5 & $1 \pm 0$ & $2 \pm 1$ & $0.01 \pm 0.01$ & $-0.34 \pm 0.02$ \\
-5 & 1 & 3 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.01 \pm 0.01$ & $-0.01 \pm 0.01$ \\
-10 & 1 & 3 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.01$ & $-0.07 \pm 0.01$ \\
-20 & 1 & 3 & 5 & $0 \pm 0$ & $0 \pm 0$ & $-0.00 \pm 0.00$ & $-0.20 \pm 0.02$ \\
-5 & 5 & 3 & 5 & $192 \pm 20$ & $38 \pm 10$ & $0.12 \pm 0.01$ & $0.01 \pm 0.01$ \\
-10 & 5 & 3 & 5 & $7 \pm 1$ & $7 \pm 1$ & $0.08 \pm 0.02$ & $-0.11 \pm 0.02$ \\
-20 & 5 & 3 & 5 & $1 \pm 0$ & $3 \pm 1$ & $0.03 \pm 0.01$ & $-0.27 \pm 0.02$ \\
-5 & 1 & 1 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.03 \pm 0.01$ & $-0.21 \pm 0.01$ \\
-10 & 1 & 1 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.27 \pm 0.01$ \\
-20 & 1 & 1 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.32 \pm 0.01$ \\
-5 & 5 & 1 & 10 & $4 \pm 2$ & $0 \pm 0$ & $0.14 \pm 0.01$ & $-0.17 \pm 0.01$ \\
-10 & 5 & 1 & 10 & $58 \pm 5$ & $43 \pm 6$ & $0.06 \pm 0.01$ & $-0.07 \pm 0.01$ \\
-20 & 5 & 1 & 10 & $6 \pm 1$ & $6 \pm 1$ & $0.03 \pm 0.01$ & $-0.24 \pm 0.01$ \\
-5 & 1 & 3 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.05 \pm 0.01$ & $-0.02 \pm 0.01$ \\
-10 & 1 & 3 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.03 \pm 0.01$ & $-0.09 \pm 0.01$ \\
-20 & 1 & 3 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.22 \pm 0.01$ \\
-5 & 5 & 3 & 10 & $45 \pm 11$ & $0 \pm 0$ & $0.26 \pm 0.01$ & $0.03 \pm 0.01$ \\
-10 & 5 & 3 & 10 & $98 \pm 10$ & $31 \pm 4$ & $0.04 \pm 0.01$ & $-0.05 \pm 0.01$ \\
-20 & 5 & 3 & 10 & $9 \pm 1$ & $8 \pm 1$ & $0.06 \pm 0.01$ & $-0.16 \pm 0.01$ \\
-5 & 1 & 1 & 20 & $1 \pm 0$ & $0 \pm 0$ & $0.05 \pm 0.01$ & $-0.24 \pm 0.01$ \\
-10 & 1 & 1 & 20 & $0 \pm 0$ & $0 \pm 0$ & $0.03 \pm 0.01$ & $-0.34 \pm 0.01$ \\
-20 & 1 & 1 & 20 & $0 \pm 0$ & $0 \pm 0$ & $0.01 \pm 0.00$ & $-0.42 \pm 0.01$ \\
-5 & 5 & 1 & 20 & $4 \pm 0$ & $1 \pm 0$ & $0.22 \pm 0.01$ & $-0.17 \pm 0.01$ \\
-10 & 5 & 1 & 20 & $9 \pm 3$ & $1 \pm 0$ & $0.07 \pm 0.01$ & $-0.24 \pm 0.01$ \\
-20 & 5 & 1 & 20 & $37 \pm 3$ & $35 \pm 3$ & $0.08 \pm 0.01$ & $-0.07 \pm 0.01$ \\
-5 & 1 & 3 & 20 & $1 \pm 0$ & $0 \pm 0$ & $0.06 \pm 0.01$ & $-0.02 \pm 0.01$ \\
-10 & 1 & 3 & 20 & $0 \pm 0$ & $0 \pm 0$ & $0.02 \pm 0.01$ & $-0.19 \pm 0.01$ \\
-20 & 1 & 3 & 20 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.01$ & $-0.31 \pm 0.01$ \\
-5 & 5 & 3 & 20 & $31 \pm 9$ & $1 \pm 0$ & $0.36 \pm 0.01$ & $0.04 \pm 0.01$ \\
-10 & 5 & 3 & 20 & $16 \pm 4$ & $1 \pm 0$ & $0.16 \pm 0.01$ & $-0.08 \pm 0.01$ \\
-20 & 5 & 3 & 20 & $48 \pm 3$ & $33 \pm 3$ & $0.06 \pm 0.01$ & $-0.07 \pm 0.01$ \\
+5 & 1 & 1 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.00$ & $-0.05 \pm 0.01$ \\
+10 & 1 & 1 & 5 & $0 \pm 0$ & $0 \pm 0$ & $-0.00 \pm 0.00$ & $-0.16 \pm 0.02$ \\
+5 & 5 & 1 & 5 & $108 \pm 11$ & $37 \pm 7$ & $0.05 \pm 0.01$ & $-0.05 \pm 0.01$ \\
+10 & 5 & 1 & 5 & $9 \pm 2$ & $5 \pm 1$ & $0.10 \pm 0.01$ & $-0.17 \pm 0.02$ \\
+5 & 1 & 3 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.01 \pm 0.01$ & $-0.01 \pm 0.00$ \\
+10 & 1 & 3 & 5 & $0 \pm 0$ & $0 \pm 0$ & $0.01 \pm 0.01$ & $-0.05 \pm 0.01$ \\
+5 & 5 & 3 & 5 & $207 \pm 27$ & $39 \pm 8$ & $0.12 \pm 0.01$ & $0.01 \pm 0.01$ \\
+10 & 5 & 3 & 5 & $8 \pm 1$ & $7 \pm 2$ & $0.11 \pm 0.02$ & $-0.10 \pm 0.02$ \\
+5 & 1 & 1 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.03 \pm 0.01$ & $-0.20 \pm 0.01$ \\
+10 & 1 & 1 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.00 \pm 0.01$ & $-0.22 \pm 0.01$ \\
+5 & 5 & 1 & 10 & $16 \pm 8$ & $0 \pm 0$ & $0.15 \pm 0.01$ & $-0.15 \pm 0.01$ \\
+10 & 5 & 1 & 10 & $59 \pm 6$ & $46 \pm 5$ & $0.06 \pm 0.01$ & $-0.07 \pm 0.01$ \\
+5 & 1 & 3 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.07 \pm 0.01$ & $0.00 \pm 0.01$ \\
+10 & 1 & 3 & 10 & $0 \pm 0$ & $0 \pm 0$ & $0.02 \pm 0.01$ & $-0.11 \pm 0.01$ \\
+5 & 5 & 3 & 10 & $49 \pm 12$ & $0 \pm 0$ & $0.25 \pm 0.01$ & $0.04 \pm 0.01$ \\
+10 & 5 & 3 & 10 & $99 \pm 9$ & $35 \pm 6$ & $0.07 \pm 0.01$ & $-0.03 \pm 0.01$ \\
 \hline
@@
 The fresh clustering has the lowest ARI, which is expected given the absence of a clear partitioning in a circular mass of observations.
-The low WCSS for the simple method is due to the proximity of the new and existing observations,
+The ability of the simple method to achieve a low WCSS is due to the proximity of the new and existing observations,
 allowing the centers to incrementally migrate into the new batch during refinement.

### 2021-12-01 simulations/newpop.R

@@ -15,10 +15,10 @@ generator <- function(npop = 10, sd = 5, holdout=1, ncells = 1000, ndim = 5) {
 set.seed(984241)
 
 settings <- expand.grid(
-    npop = c(5, 10, 20),
+    npop = c(5, 10),
     sd = c(1, 5),
     holdout = c(1, 3),
-    k = c(5, 10, 20)
+    k = c(5, 10)
 )
 
 library(KmeansReinitialization2021)
